Searching for circuit complexity lower bounds using SAT solvers

### 2022-01-01 report/draft1.tex
continuing report

*mention valiant universal circuits?

@@

We also present a new reduction *how to cite*

*Extension axioms - let’s actually try them first

@@

Denote by \(B_{n,m}\) the set of all Boolean functions \(f: \{0,1\}^n \to \{0,1\}^m\) and let \(B_n = B_{n,1}\). A circuit over the basis \(A \subseteq B_2\) is a directed acyclic graph with nodes of in-degree 0 or 2. Nodes of in-degree 0 are called inputs. Nodes of in-degree 2 are assigned functions from \(A\) and are called gates. Some gates may be also be marked as outputs.

@@

In this paper, we mainly consider circuits over the basis \(C_2 = \{\lor, \land, \neg\}\), where \(\neg\) denotes the Boolean function which outputs the negation of its first argument.

@@

@@

\subsubsection{third reduction}

The third is due to Jan Pich *HOW 2 CITE THO* and also requires a circuit over the basis \(C_2\).

Include the following variables:

\begin{enumerate}
  \item \(e^i_a\) \((n \leq i < N, a\in\{0,1\}^n)\) is the output of the \(i\)-th gate when the input to the circuit is \(a\). It gives \(O(N\cdot 2^n)\) variables.
  \item \(x_{i,j}\) \(( 0 \leq i < N+n,0 \leq j < i)\) is true iff j is a predecessor of i. It gives \(O(N\cdot N+n)\) variables.
  \item \(g^0_i, g^1_i\) \((n \leq i < N)\) together encode the function assigned to the \(i\)-th gate by their truth values, giving \(2N\) variables:

@@

@@

\subsection{Possible extensions}

>>>>

*Mention valiant universal circuits?

*Should look for other things

@@

We also present a new reduction *not sure what to say about this*

*Extension axioms - should actually try them first

@@

Denote by \(B_{n,m}\) the set of all Boolean functions \(f: \{0,1\}^n \to \{0,1\}^m\) and let \(B_n = B_{n,1}\). A circuit over the basis \(A \subseteq B_2\) is a directed acyclic graph with nodes of in-degree 0 or 2. Nodes of in-degree 0 are called inputs. Nodes of in-degree 2 are assigned functions from \(A\) and are called gates. Some nodes may be also be marked as outputs.

@@

In this paper, we mainly consider circuits over the De Morgan basis \[C_2 = \{\lor, \land, \neg\}\] where \(\neg\) denotes the Boolean function which outputs the negation of its first argument.

@@

\subsubsection{Existing reductions}

@@

*Also am bothered that this is the only reduction that allows constants in the circuit - it wouldn't be hard to remove that aspect.*

\subsubsection{A new reduction}

The third *Don't know how to credit Jan* also requires a circuit over the basis \(C_2\).

It uses the following variables:

\begin{enumerate}
  \item \(e^i_a\) \((n \leq i < N, a\in\{0,1\}^n)\) is the value of the \(i\)-th gate when the input to the circuit is \(a\). It gives \(O(N\cdot 2^n)\) variables.
  \item \(x_{i,j}\) \(( 0 \leq i < N+n,0 \leq j < i)\) is true iff j is a predecessor of i. It gives \(O(N\cdot N+n)\) variables.
  \item \(g^0_i, g^1_i\) \((n \leq i < N)\) together encode the function assigned to the \(i\)-th gate by their truth values, giving \(2N\) variables:

@@

We encode the requirements for the circuit by the following clauses:

\begin{enumerate}
  \item For all \((i,j,k)\) with \(n \leq i < n+N\), \(0 \leq j < i\) and \(j < k < i\), and for all \(a\in\{0,1\}^n\), if \(j\) and \(k\) are both predecessors of \(i\), then \(e^i_a\) must take the value computed by the \(i\)th gate's assigned function on \(e^j_a\) and \(e^k_a\). A clause must be added for each function in the basis. For example, the clause for when the \(i\)th gate is an \(\land\)-gate:

\[
(g^0_i\wedge \neg g^1_i)\wedge 
(x_{i,j}\wedge x_{i,k})\rightarrow 
(e^i_a\leftrightarrow e^j_a\wedge e^k_a)
\]

  (Note that the gate must compute the conjunction of any two predecessors' variables, so it can have a maximum of two predecessors.) This leads to \(O(2^n\cdot N^3)\) 7-clauses.

  \item For all \(i\) with \(n \leq i < n+N\), gate \(i\) must have at least one predecessor. This leads to \(O(N)\) \(O(N)\)-clauses.

  \item For all \((i,j)\) with \(n \leq i < n+N\), \(0 \leq j < i\), if \(i\) is an \(\lor\)-gate or \(\land\)-gate with \(j\) its predecessor, it must have another distinct predecessor (i.e. it must have at least 2 overall). For example, the clause for when the \(i\)th gate is an \(\land\)-gate:

\[
(g^0_i \land \neg g^1_i) \rightarrow \left( x_{i,j} \rightarrow \bigvee_{j<k<i} x_{i,k} \right)
\]

  This leads to \(O(N^2)\) \(O(N)\)-clauses.

  \item For all \(i\) with \(0 \leq i < n\), \(a\in\{0,1\}^n\), the \(i\)-th input must equal \(a_i\). This leads to \(O(n \cdot 2^n)\) 1-clauses.

  \item For all \(a\in\{0,1\}^n\), the final m gates (i.e. the outputs) should agree with the output values given for \(a\) in the target truth table. This leads to \(O(2^n)\) 1-clauses.

@@

\subsection{Further directions}
